Dense depth estimation from image pairs

### 2021-06-30 paper/paper.tex
stuff

\newcommand{\Part}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\Partt}[2]{\frac{\partial^2 #1}{\partial {#2}^2}}
\newcommand{\Partts}[3]{\frac{\partial^2 #1}{\partial {#2} \partial {#3}}}

@@

\section{Dense variational methods}
Here we give a short presentation of early variational methods used in computer vision.
The methods outlined here serve as the algorithmic lineage of the method of Cremers et. al. \cite{dense_geometry}.
See the TUM lecture series by Cremers \cite{variational_lectures}, freely available online, for detailed descriptions of the mathematical underpinnings
and fundamental algorithms.

\subsection{The Horn-Schunck method for dense optical flow estimation with a regularized variational minimization.}
A well-known 1981 paper on optical flow estimation by Horn and Schunck \cite{horn_schunck} introduced what is now called the ``Horn-Schunck method'', one of the
first widespread variational algorithms used by the computer vision community
(See \cite{szeliski} for an introduction to the problem of optical flow estimation.) The Horn-Schunck method, when introduced, was unique in that
it computes a \textit{dense} field of motion vectors. The method is usefully formulated as a continuous optimization problem.
Given a stream of grayscale images $I^t : [0,1]^2 \rightarrow \mathbb{R}$, we would like to compute, for each point in the image at time $t$, a (space and time) velocity vector $(u,v,1)$
which which minimizes the directional derivative
    $$\nabla I^t \cdot(u, v, 1) = \Part{I^t}{x} u + \Part{I^t}{y}v + \Part{I^t}{t}.$$
This effectively finds the most likely ``motion vector'' which connects this point, as time goes in, to a point
in a subsequent image with the closest intensity value.
We can then formulate the functional
\begin{align*}
    E(u, v) = \int_0^1\int_0^1 \Part{I^t}{x} u(x,y) + \Part{I^t}{y}v(x,y) + \Part{I^t}{t}(x,y) \,dx\,dy,
\end{align*}
and our algorithm is simply the optimization problem
\begin{equation}\label{horn_schunck_unregularized}
    \min_{u,v}E(u, v).
\end{equation}
At this point, as in all variational optimization methods,
the essential conceptual work is done, and the rest of the work is in the choice of optimization strategy,
and our method of discretization. However, there is a fatal flaw in \eqref{horn_schunck_unregularized} ---
given

\subsection{Total-variation denoising with the Rudin-Osher-Fatemi (ROF) model}
The problem of image denoising can be put in a global optimization
framework. A cost function is formulated that penalizes ``noise'' and rewards closeness to the original (noisy) image. The algorithm then
consists of minimizing this cost function over all possible candidate ``denoised'' images.

@@

A common finite difference approximation for $\nabla\hat{I}(x,y)$, valid for non-boundary pixels, is
\begin{equation}
    \hat{\nabla}I_{i,j} = \left(\frac{I_{i+1,j} - I_{i-1,j}}{2\Delta x}, \frac{I_{i,j+1} - I_{i,j-1}}{2\Delta y}\right)^T,
\end{equation}
where $\Delta x, \Delta y$ are pixel extents in the image domain. Where the original image is discontinuous, such as at edges,
this finite difference vector can be very large. Furthermore, the set of pixels whose finite-difference stencils extend over discontinuities is
\textit{not} neglible in the finite approximation of integral \eqref{rof_continuous}. A quadratic regularizer, such as
$\frac{\lambda}{2} \|\nabla \hat{I}\|^2$ will harshly penalize the appearance of sharp discontinuities, since a large value returned
by a finite difference is squared. The main idea behind the ROF model is to use the non-squared ``total-variation'' $\lambda \|\nabla \hat{I}\|$.
While this is notably more difficult to optimize (precluding the use of simple linear least squares), the final effect is a preservation
of isolated discontinuities such as edges and stripe patterns, while still penalizing large patches of interior noise. Notably, the ROF model can be solved by a non-linear diffusion process, performing gradient descent to solve the Euler-Lagrange
equations of the cost functional. See their classic paper \cite{rof} for details, and \cite{cremers_rof} for a more recent discussion.

@@

\subsection{Total-variation for dense optical flow estimation}
In \cite{tv_optical_flow}, Zach et al. formulate the optical flow problem (see \cite{szeliski} for an introduction to this topic).
as a global optimization with non-quadratic data and regularizer terms. This gives a modification
of the Horn-Schunck algorithm which is more robust to outliers and which tends to preserve motion discontinuities across object boundaries.
Zach et. al's main contribution is their method of optimization,
which we will reproduce in the context of dense depth map estimation.
See their paper \cite{tv_optical_flow} for full details in the context of dense optical flow estimation.

\section{Total variation for depth map estimation from image pairs}
The paper by Cremer's et al. \cite{dense_geometry} applies the method of Zach et al. \cite{tv_optical_flow} to
the problem of dense depth-map
reconstruction from collections of (grayscale) images. For simplicity, we restrict our attention to the case of two images only ---
see \cite{dense_geometry} for details on the generalization. We assume (as in \cite{dense_geometry}) that we have an accurate (and constant) intrinsic camera matrix,
our images are undistorted, and, importantly, that we have accurate camera pose estimations.
Our fully non-linear cost function is
\begin{equation}
\begin{split}
E(h) &= \lambda \int_{\Omega_0}\left\|I_1(\pi(\exp(\hat{\eta}_1)X(x,h)) - I_0(\pi(x))\right\|
     + \|\nabla h\| d^2 x,
\end{split}
\end{equation}
which has TV-$L^1$ \cite{tv_optical_flow} data and regularizer terms.
(---REDO reprojection notation)
Let $\rho(h) = \left\|I_1(\pi(\exp(\hat{\eta}_1)X(x,h)) - I_0(\pi(x))\right\|$. Since the data term is non-quadratic, we
cannot apply the ROF optimization method directly. However, we can introduce
an auxiliary depth map $h^\prime$ to separate the data and regularizer terms,
and introduce a penalizer for disparity between $h^\prime$ and $h$. The
new separable cost function is
\begin{equation}\label{separable_cost}
    \hat{E}_\theta(h, \hat{h}) = \int_\Omega \lambda \|\rho(h)\| + \frac{1}{2\theta}(h - \hat{h})^2 + \|\nabla \hat{h}\|.
\end{equation}
The parameter $\theta$ is set to some small constant.
The cost function $E_\theta$ is separated into functions of $h$ and $h^\prime$,
and $h$ and $h^\prime$ are alternately optimized in the following way:
\begin{itemize}
    \item For $h$ fixed, solve
        \begin{equation}\label{minh}
            \min_{h^\prime} \int_\Omega \lambda \|\nabla h^\prime\| + \frac{1}{2\theta}(h - h^\prime)^2\,dx
        \end{equation}
    using the ROF optimization methods.
    \item For $h^\prime$ fixed, solve
        \begin{equation}\label{minhhat}
            \min_h \int_\Omega \|\rho(h)\| + \frac{1}{2\theta}(h - h^\prime)^2\,dx.
        \end{equation}
        Since $\rho(h)$ measures point-wise reprojection error, this
        optimization can be done point-wise.
\end{itemize}
The ROF cost function \eqref{minh} can be minimized using the iterative method of Chambolle \cite{chambolle}. For completeness, we reproduce this
method here in the context of dense depth map estimation. We provide no proofs --- see \cite{chambolle} for full details.

@@

\begin{figure}[htbp]
\begin{subfigure}[b]{0.5\textwidth}
\centerline{\includegraphics[width=\textwidth]{figures/noisevis_groundtruth.png}}
\caption{Ground truth depth data}
\label{noisevis_groundtruth}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\centerline{\includegraphics[width=\textwidth]{figures/noisevis_computed.png}}
\caption{Brute-force computed, with no regularizer}
\label{noisevis_computed}
\end{subfigure}
\end{figure}

@@

>>>>

\newcommand{\Part}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\Partt}[2]{\frac{\partial^2 #1}{\partial {#2}^2}}
\newcommand{\Partts}[3]{\frac{\partial^2 #1}{\partial {#2} \partial {#3}}}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

@@

\section{Dense variational methods}
Here we give a short presentation of early variational methods used in computer vision ---
the Horn-Schunck and Rudin-Osher-Fatemi algorithms --- and a more recent related method of Zach et al. \cite{tv_optical_flow}
The methods outlined here serve as the algorithmic lineage of the method of Cremers et. al. \cite{dense_geometry}.
See the TUM lecture series by Cremers \cite{variational_lectures}, freely available online, for detailed descriptions of the mathematical underpinnings
and fundamental algorithms.

\subsection{The Horn-Schunck method for dense optical flow estimation with a regularized variational minimization.}
A well-known 1981 paper on optical flow estimation by Horn and Schunck \cite{horn_schunck} introduced what is now called the ``Horn-Schunck method'', one of the
first widespread variational algorithms used by the computer vision community
(See \cite{szeliski} for an introduction to the problem of optical flow estimation.) The Horn-Schunck method, when introduced, was unique in that
it computes a \textit{dense} field of motion vectors. The method is usefully formulated as a continuous optimization problem.
Given a stream of grayscale images $I^t : [0,1]^2 \rightarrow \mathbb{R}$, we would like to compute, for each point in the image at time $t$, a (space and time) velocity vector $(u,v,1)$
which which minimizes the directional derivative
    $$\nabla I^t \cdot(u, v, 1) = \Part{I^t}{x} u + \Part{I^t}{y}v + \Part{I^t}{t}.$$
This effectively finds the most likely ``motion vector'' which connects this point, as time goes on, to a point
in a subsequent image with the closest intensity value.
We can then formulate the functional
\begin{equation}
    E(u, v) = \int_0^1\int_0^1\left(\Part{I^t}{x} u(x,y) + \Part{I^t}{y}v(x,y) + \Part{I^t}{t}(x,y)\right)^2 \,dx\,dy,
\end{equation}
and our algorithm is simply the optimization problem
\begin{equation}\label{horn_schunck_unregularized}
    \min_{u,v}E(u, v).
\end{equation}
At this point, as in all variational optimization methods,
the essential conceptual work is done, and the rest of the work is in the choice of optimization strategy,
and our method of discretization. However, there is a fatal flaw in \eqref{horn_schunck_unregularized}.
Suppose that the brightness constancy assumption is fully satisfied; that is, we have $\nabla I^t \cdot (u, v, 1) = 0$
for ground truth motion vector $(u,v,1)$. We immediately see that this is one linear equation in two variables,
    $$u\Part{I^t}{x} + v\Part{I^t}{y} = -\Part{I^t}{t},$$
and that for any $u^*,v^*$ such that
    $$u^*\Part{I^t}{x} + v^*\Part{I^t}{y} = 0,$$
we have another solution, $(u + u^*, v + v^*)$ (see figure \ref{aperture_problem}). This is called the \textit{aperture problem} \cite{aperture_problem}, and is due to
the fact that the velocity vector field is estimated locally. The key modification of Horn and Schunck is the addition of a regularizer term which penalizes
high variation of the motion field, based on the observation that, except on object boundaries and occlusions, motion vector fields should be smooth. Their modified algorithm
solves the optimization problem
\begin{equation}\label{horn_schunck_regularized}
    \min_{u,v}\left\{E(u, v) + R(u,v)\right\},
\end{equation}
where $R(u,v)$ is the regularizer term
    $$\alpha^2\int_0^1\int_0^1\norm{u}^2 + \norm{v}^2\,dx\,dy,$$
for regularizer parameter $\alpha$ (the higher $\alpha$, the smoother the resulting vector field).
Quadratic data and regularizer terms are chosen in order to reduce the discretized Euler-Lagrange equations to a system of linear equations solved for
the global minimizer (see \cite{variational_calculus} and \cite{variational_lectures} for more information on variational calculus).
A quadratic regularizer, however, disproportionately penalizes discontinuities, such as across real object boundaries. We will
address this problem below.

\subsection{The Rudin-Osher-Fatemi (ROF) model for image denoising with total variation}
The problem of image denoising can be put in a global optimization
framework. We care here about the form of the optimization problem, rather than the application, and the ROF model turns out to be exactly
the same optimization required in the ``regularization step'' in \cite{tv_optical_flow}, \cite{dense_geometry}.
A cost function is formulated that penalizes ``noise'' and rewards closeness to the original (noisy) image. The algorithm then
consists of minimizing this cost function over all possible candidate ``denoised'' images. In fact, this algorithm is effectively the same
as the above Horn-Schunck method, utilizing a slightly different data term, with a more robust regularizer.

@@

A common finite difference approximation for $\nabla\hat{I}(x,y)$, valid for non-boundary pixels, is
\begin{equation}
    \hat{\nabla}I_{i,j} = \left(\frac{I_{i+1,j} - I_{i-1,j}}{2\Delta x}, \frac{I_{i,j+1} - I_{i,j-1}}{2\Delta y}\right)^T,
\end{equation}
where $\Delta x, \Delta y$ are pixel extents in the image domain. Where the original image is discontinuous, such as at edges,
this finite difference vector can be very large. Furthermore, the set of pixels whose finite-difference stencils extend over discontinuities is
\textit{not} neglible in the finite approximation of integral \eqref{rof_continuous}. A quadratic regularizer, such as
$\frac{\lambda}{2} \|\nabla \hat{I}\|^2$ will harshly penalize the appearance of sharp discontinuities, since a large value returned
by a finite difference is squared. The main idea behind the ROF model is to use the non-squared ``total-variation'' $\lambda \|\nabla \hat{I}\|$.
While this is notably more difficult to optimize (precluding the use of simple linear least squares), the final effect is a preservation
of isolated discontinuities such as edges and stripe patterns, while still penalizing large patches of interior noise. Notably, the ROF model can be solved by a non-linear diffusion process, performing gradient descent to solve the Euler-Lagrange
equations of the cost functional \cite{variational_calculus}. See their classic paper \cite{rof} for details, and \cite{variational_lectures} for a more recent discussion.

@@

\subsection{Total-variation for dense optical flow estimation}
In \cite{tv_optical_flow}, Zach et al. formulate the optical flow problem
as a global optimization with non-quadratic data and regularizer terms.
This gives a modification
of the Horn-Schunck algorithm which is more robust to outliers and which tends to preserve motion discontinuities across object boundaries.
Zach et al.'s main contribution is their method of optimization,
which we will reproduce in the context of dense depth map estimation.
See their paper \cite{tv_optical_flow} for full details in the context of dense optical flow estimation.

\section{Total variation for depth map estimation from image pairs}
The paper by Cremer's et al. \cite{dense_geometry} applies the method of Zach et al. \cite{tv_optical_flow} to
the problem of dense depth-map
reconstruction from collections of (grayscale) images. For simplicity, we restrict our attention to the case of two images only ---
see \cite{dense_geometry} for details on the generalization. We assume (as in \cite{dense_geometry}) that we have an accurate (and constant) intrinsic camera matrix,
our images are undistorted, and, importantly, that we have accurate camera pose estimations. We describe here a slightly modified version of Cremer's method.

Let $C_1$ and $C_2$ denote the two cameras, $I_1$ and $I_2$ denote the two images, and let $h$ be the estimated depth map for $C_1$.
Our fully non-linear cost function is
\begin{equation}
\begin{split}
% E(h) &= \lambda \int_0^1\int_0^1{\Omega_0}\left\|I_1(\pi(\exp(\hat{\eta}_1)X(x,h)) - I_0(\pi(x))\right\|
%      + \|\nabla h\| \,dx\,dy,
E(h) &= \lambda \frac{1}{|\Omega^*|}\int_{\Omega^*}\left\|I_2(\text{reproj}(x,y)) - I_1(x,y)\right\|\,dx\,dy \\
     &+ \int_0^1\int_0^1\|\nabla h\| \,dx\,dy.
\end{split}
\end{equation}
which has TV-$L^1$ \cite{tv_optical_flow} data and regularizer terms. $\text{reproj}(x,y)$ is where the non-linear complexity is.
$x$ and $y$ denote image coordinates in $I_1$, and $h$ gives a transformation from these image coordinates to world-space,
given the camera pose of $C_1$. We can then project this point to the image space of $I_2$. $\Omega^* \subset [0,1]^2$
is defined as
\begin{equation}
    \Omega_0 \coloneqq \left\{(x,y) \in [0,1]^2 \mid \text{reproj}(x,y) \in [0,1]^2\right\}.
\end{equation}
Therefore, our data term is weighted over those pixels which have valid reprojections. Without the $1/|\Omega^*|$ factor
in the data term, our optimization would likely cause $h$ to be a constant function (giving zero regularizer cost), large enough such that
all reprojections are invalid (giving zero data cost) --- Weighting by $1/|\Omega^*|$ has been effective in our implementation.

Let $\rho(h) = \left\|I_1(\pi(\exp(\hat{\eta}_1)X(x,h)) - I_0(\pi(x))\right\|$. Since the data term is non-quadratic, we
cannot apply the ROF optimization method directly. However, we can introduce
an auxiliary depth map $h^\prime$ to separate the data and regularizer terms,
and introduce a penalizer for disparity between $h^\prime$ and $h$. The
new separable cost function is
\begin{equation}\label{separable_cost}
    \hat{E}_\theta(h, \hat{h}) = \int_\Omega \lambda \|\rho(h)\| + \frac{1}{2\theta}(h - \hat{h})^2 + \|\nabla \hat{h}\|.
\end{equation}
The parameter $\theta$ is set to some small constant.
The cost function $E_\theta$ is separated into functions of $h$ and $h^\prime$,
and $h$ and $h^\prime$ are alternately optimized in the following way:
\begin{itemize}
    \item For $h$ fixed, solve
        \begin{equation}\label{minh}
            \min_{h^\prime} \int_\Omega \lambda \|\nabla h^\prime\| + \frac{1}{2\theta}(h - h^\prime)^2\,dx
        \end{equation}
    using the ROF optimization methods.
    \item For $h^\prime$ fixed, solve
        \begin{equation}\label{minhhat}
            \min_h \int_\Omega \|\rho(h)\| + \frac{1}{2\theta}(h - h^\prime)^2\,dx.
        \end{equation}
        Since $\rho(h)$ measures point-wise reprojection error, this
        optimization can be done point-wise.
\end{itemize}
The ROF cost function \eqref{minh} can be minimized using the iterative method of Chambolle \cite{chambolle}. For completeness, we reproduce this
method here in the context of dense depth map estimation. We provide no proofs --- see \cite{chambolle} for full details.

@@

\begin{figure}[htbp]
\begin{subfigure}[b]{0.5\textwidth}
\centerline{\includegraphics[width=\textwidth]{figures/noisevis_groundtruth_text.png}}
\caption{Ground truth depth data}
\label{noisevis_groundtruth}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\centerline{\includegraphics[width=\textwidth]{figures/noisevis_computed_text.png}}
\caption{Brute-force computed, with no regularizer}
\label{noisevis_computed}
\end{subfigure}
\end{figure}

@@

\bibitem{variational_calculus}
    Gelfand, I. M., \& Fomin, S. V. (2000). Calculus of Variations (Dover Books on Mathematics). Dover Publications.
